---
title: "R functions"
format: html
editor: visual
---

# 4.1 Characteristics of R functions

```{r}
seq(from = 5, to = 8, by = 0.5)
```

```{r}
seq(5, 8, 0.5)
```

```{r}
seq(5, 8, length.out = 26)  # 26 numbers in the sequence
```

```{r}
seq(5, 8)
```

```{r}
log(base=10)
```

```{r}
log(15)  # R uses the default value of `exp(1)`
```

```{r}
log(15, base = 10)  # R uses our value 10
```

```{r}
log(3)

log(x=3)
log(x=3, exp(1))
log(x=3, base=exp(1))

log(3, exp(1))
log(3, base=exp(1))

log(base=exp(1), 3)
log(base=exp(1), x=3)
```

```{r}
log(exp(1), 3)
```

```{r}
date()
```

# 4.2 Mathematical functions

## 4.2.1 Logarithmic and exponential functions

```{r}
log(100)     # natural logarithm (or base-e logarithm)
log(0.05)
log2(100)    # base-2 logarithm
log10(100)   # base-10 logarithm
```

```{r}
exp(5)       # exponential exp(1)^5
exp(0.5)     # exponential exp(1)^(1/2)
```

## 4.2.2 Trigonometric functions

```{r}
sin(pi/2)  # pi approximately equals to 3.14
cos(pi)
tan(pi/3)
```

## 4.2.3 Other mathematical functions

```{r}
sqrt(9)       # returns the squared root of a number
abs(-9)       # returns absolute value of a number
sign(-9)      # returns the sign of a number, -1, 0, or 1
factorial(3)  # factorial 3! = 1x2x3
choose(6, 2)  # number of combinations without replacement 6!/(6-2)!2!
```

## 4.2.4 The round() function

```{r}
round(7 / 3)  # rounding 7/3 (2.333) to the nearest integer
round(7 / 3, digits = 2)  # rounding 7/3 to two decimal places
```

If the first digit that is dropped is exactly 5, R uses a rule that’s common in programming languages: Always round to the nearest **even** number

```{r}
round(1.5)
round(2.5)
round(4.5)
round(5.5)
```

```{r}
ceiling(16.2)       # round to the nearest integer above
floor(16.2)         # round to the nearest integer below
trunc(125.2395)     #  truncates the values in the decimal places
signif (2718214, 3)  # round to the specified number of significant digits
```

# 4.3 The sessionInfo() and option() functions

```{r}
sessionInfo()
```

```{r}
options()
```

```{r}
help(options)
```

# 4.4 User defined functions

```{r}
log7 <- function(x) {
  log(x, base = 7)
  }

# calculate the logarithm of 5 to base 7
log7(5)
```
