---
title: "R objects"
format: html
editor: visual
---

# 6.1 What are the objects in R

Objects in R have many properties associated with them, called **attributes**.

These properties explain what an object represents and how it should be interpreted by R.

Two of the most important attributes of an R object are the `class` and the `dimension` :

```{r}
class(iris)
```

```{r}
dim(iris)
```

```{r}
attributes(iris)
```

# 6.2 Named storage of objects

## 6.2.1 Assignment operator (`<-`)

```{r}
x <- 1/40
x
```

```{r}
(x <- 1/40)
```

```{r}
log(x)
```

Using spaces before and after comparison operators is recommended :

```{r}
x < -1/50
```

```{r}
x<-1/50
x
```

## 6.2.2 Other types of assignment

```{r}
x = 1/40
x
```

```{r}
1/40 -> x
x
```

# 6.3 Reassigning an object

```{r}
x
```

```{r}
x <- 100
x
```

```{r}
x <- x + 1
x
```

# 6.4 Legal object names

Object names must start with a letter and can contain letters, numbers, underscores and periods.

They cannot start with a number or underscore, nor contain spaces at all.

They cannot contain reserved words.

```{r}
make.names(names = "test-1&2")
```

```{r}
Y <- 50
Y
y
```

# 6.5 We are not limited to store numbers in objects

```{r}
sentence <- "the cat sat on the mat"
```

```{r}
sentence + 1
```
