---
title: "Basic concepts of probability"
format: html
editor: visual
---

# 13.1 Packages we need

```{r}
# dice graphs
library(tidydice)
library(ggplot2)
library(ggpolypath)

# venn diagrams
library(ggvenn)
library(venn)
library(TeachingDemos)
```

# 13.2 Sample space and random events

-   A deterministic phenomenon always produces the same outcome each time it is repeated under the same conditions

-   A random phenomenon is characterized by conditions under which the result cannot be determined with certainty before it occurs (e.g. coin toss -\> H or T)

-   The sample space is defined as the set of all possible outcomes of a random experiment

    -   6-sided die: sample space is {1,2,3,4,5,6}

```{r}
force_dice(1:6) |> 
  plot_dice(detailed = TRUE, fill_success = "white") + 
  theme(plot.title = element_blank())
```

-   A random event is a sub-set of sample space, including a number of possible outcomes of the experiment

    -   event "even number" : A = {2,4,6}

    -   event "odd number" : B = {1,3,5}

-   Simple event: an event that consists of a single outcome from the sample space

-   Compound event: an event that consists of more than a single outcome from the sample space

# 13.3 Operations of events using set theory and Venn diagrams

## 13.3.1 Union of events: A U B

Collection of all outcomes that are in A or in B or in both

```{r}
A <- c(2, 4, 6)      # A = {2, 4, 6}
B <- c(1, 2)         # B = {1, 2}

union(A, B)          # A∪B = {2, 4, 6, 1} 
```

```{r}
venn("A + B", zcolor = "#7F7FFF", opacity = 1, ggplot = TRUE) +
  annotate("text", x = 35, y = 950, label = "Ω") +
  annotate("text", x = 500, y = 750, label = "AUB")
```

## 13.3.2 Intersection of events

All outcomes that are in both A and B

```{r}
# A = {2, 4, 6}
# B = {1, 2}

intersect(A, B)
```

```{r}
venn("A B", zcolor = "#7F7FFF", opacity = 1,  ggplot = TRUE) +
  annotate("text", x = 35, y = 950, label = "Ω") +
  annotate("text", x = 500, y = 530, label = "A∩B")
```

## 13.3.3 Complement events

All outcomes of the sample space that don't belong to A U B

```{r}
# A = {2, 4, 6}
# B = {1, 2}

AUB <- union(A, B)                     # A∪B = {2, 4, 6, 1} 


sample_space <- c(1, 2, 3, 4, 5, 6)    # sample_space = {1, 2, 3, 4, 5, 6}

setdiff(sample_space, AUB)
```

```{r}
venn("A + B", zcolor = "white", opacity = 1, box = FALSE, ggplot = TRUE) +
theme(panel.background = element_rect(fill = "#7F7FFF")) +
  annotate("text", x = 5, y = 1000, label = "Ω") +
  annotate("text", x = 500, y = 950, label = "(AUB)^{C}", parse =TRUE)
```

## 13.3.4 Mutually exclusive events

A = "the number rolled is even"

B = "the number rolled is odd"

```{r}
A <- c(2, 4, 6)      
C <- c(1, 3, 5)         

intersect(A, C)         
```

```{r}
# List of items
x2 <- list("A" = A, "C" = C)

ggvenn(x2, fill_alpha = 1, auto_scale = T, show_percentage = FALSE) +
  labs(title = "Ω") +
  scale_fill_manual(values = c("#7F7FFF", "#7F7FFF")) +
  theme(plot.title = element_text(size = 20),
        plot.background = element_rect(fill = "white"))
```

# 13.4 Probability

The probability of an event should fulfill three axioms defined by Kolmogorov:

1.  The probability of an event A is a non-negative number, $P(A) \ge 0$
2.  The probability of all possible outcomes, or sample space $P(\Omega)=1$
3.  If A and B are two mutually exclusive events (also known as disjoint events), then $P(A \cup B) = P(A)\ +\ P(B)$ and $P(A\cap B)=0$

## 13.4.1 Definition of probability

List out all the equally probable outcomes of an experiment and determine how many of them are favorable for the event A to occur:

$$
P(A)=\frac{Number\ of\ outcomes\ favourable\ to\ the\ event\ A}{Total\ number\ of\ possible\ outcomes}
$$

What is the theoretical probability of rolling the number 5 when we roll a six-sided fair die once?

$$
P(rolling\ 5)=\frac{1\ outcome\ favourable\ to\ the\ event}{6\ possible\ outcomes}
$$

```{r}
dice_tbl <- force_dice(1:6, success = 5)
dice_tbl
```

```{r}
sum(dice_tbl$success)/length(dice_tbl$result)
```

What's the probability of rolling either a 5 or a 6?

$$
P(rolling\ a\ 5\ OR\ a\ 6)=\frac{2\ outcomes\ favourable\ to\ the\ event}{6\ possible\ outcomes}=\frac{2}{6}=\frac{1}{3}\approx0.33
$$

```{r}
dice_tbl <- force_dice(1:6, success = c(5, 6))
dice_tbl
```

```{r}
sum(dice_tbl$success)/length(dice_tbl$result)
```

The experimental probability is based on data from repetitions of the same experiment.

$$
P(A)=\frac{number\ of\ times\ A\ occured}{total\ number\ of\ experiments}
$$

```{r}
set.seed(348)
roll_dice(times = 100) |>  
  ggplot(aes(x = result)) +
  geom_bar(fill = "#0071BF", width = 0.65) +
  geom_text(aes(label=after_stat(count)), stat = "count", 
            vjust = 1.5, colour = "white") +
  scale_x_continuous(breaks = c(1:6), labels = factor(1:6)) +
  theme_minimal(base_size = 14)
```

$$
P(rolling\ a\ 5)=\frac{20\ times\ the\ number\ 5\ occured}{100\ experiments}=\frac{20}{100}=0.20\ or\ 20\%
$$

```{r}
set.seed(128)
roll_dice(times = 10000) |>  
  ggplot(aes(x = result)) +
  geom_bar(fill = "#0071BF", width = 0.65) +
  geom_text(aes(label=after_stat(count)), stat = "count", 
            vjust = 1.5, colour = "white") +
  scale_x_continuous(breaks = c(1:6), labels = factor(1:6)) +
  theme_minimal(base_size = 14)
```

$$
P(rolling\ a\ 5)=\frac{1684\ times\ the\ number\ 5\ occured}{10000\ experiments}=\frac{1684}{10000}=0.1684\ or\ 16.84\%
$$

Law of large numbers: the more times the experiment is performed, the closer the experimental probability approaches the theoretical probability.

"Subjective" probability (Bayesian approach) : the probability assigned to an event represents the degree of belief that the event will occur in a given try of the experiment, and it implies an element of subjectivity.

Fundamental properties of probability:

1.  The probability of the null event is zero, $P(\emptyset) = 0$
2.  The probability of the complement event A satisfies the property:

$$P(A') = 1-P(A)$$

3.  The probability of the union of two events satisfies the general property that (addition rule of probability):

    $$P(A\cup B) = P(A) + P(B) - P(A\cap B)$$

## 13.4.2 The conditional probability

The outcome of event A depends on the outcome of event B:

$$
P(A\cap B) = P(A|B).P(B)
$$

or

$$
P(A|B)=\frac{P(A\cap B)}{P(B)}
$$

We roll two fair six-sided dice. What is the probability that the first roll is a 3, given that the sum of two rolls is 8?

$$
\Omega=\{(1,1),(1,2), ..., (1,6), (2,1), ..., (6,6)\}
$$

A = the first roll shows 3, and the second any number

B = the sum of two rolls is 8

$$P(A)=\frac {6}{36} =\frac{1}{6}$$

$$
B=\{(2,6),(3,5), (4,4), (5,3), (6,2)\}
$$

$$
P(B)=\frac {5}{36}
$$ $$
A\cap B=\{(3,5)\}
$$

$$
P(A\cap B)=\frac {1}{36}
$$

$$
P(A|B)=\frac{P(A\cap B)}{P(B)}=\frac{\frac{1}{36}}{\frac{5}{36}} = \frac{1}{5}
$$

## 13.4.3 Bayes' theorem

$$
P(A\cap B) = P(A|B).P(B)
$$

and $$
P(A\cap B) = P(B|A).P(A)
$$

$$
P(A|B)=\frac{P(A\cap B)}{P(B)}
$$

so

$$
P(A|B)=\frac{P(B|A).P(A)}{P(B)}
$$

We are interested in calculating the probability of developing lung cancer if a person smokes tobacco for a long time, $P(Cancer|Smoker)$

8% of the population has lung cancer: $P(Cancer)=0.08$

30% of the population are chronic smokers: $P(Smoker)=0.30$

60% of all people who have lung cancer are smokers: $P(Smoker|Cancer)=0.6$

$$
P(Cancer|Smoker)=\frac{P(Smoker|Cancer).P(Cancer)}{P(Smoker)}=\frac{0.6*0.08}{0.3}=\frac{0.048}{3}=0.016
$$

## 13.4.4 Independence of events

If A and B are independent, then the conditional probability is $P(A|B)=P(A)$, i.e. the occurence of event B has no influence on the occurence of A and $P(B|A)=P(B)$, i.e. the occurence of event A has no influence on the occurence of B.

Two events A and B are said to be independent if:

$$
P(A\cap B)=P(A).P(B)
$$

This is known as the multiplication rule of probability.
