---
title: "Basic concepts of probability"
format: html
editor: visual
---

# 13.1 Packages we need

```{r}
# dice graphs
library(tidydice)
library(ggplot2)
library(ggpolypath)

# venn diagrams
library(ggvenn)
library(venn)
library(TeachingDemos)
```

# 13.2 Sample space and random events

-   A deterministic phenomenon always produces the same outcome each time it is repeated under the same conditions

-   A random phenomenon is characterized by conditions under which the result cannot be determined with certainty before it occurs (e.g. coin toss -\> H or T)

-   The sample space is defined as the set of all possible outcomes of a random experiment

    -   6-sided die: sample space is {1,2,3,4,5,6}

```{r}
force_dice(1:6) |> 
  plot_dice(detailed = TRUE, fill_success = "white") + 
  theme(plot.title = element_blank())
```

-   A random event is a sub-set of sample space, including a number of possible outcomes of the experiment

    -   event "even number" : A = {2,4,6}

    -   event "odd number" : B = {1,3,5}

-   Simple event: an event that consists of a single outcome from the sample space

-   Compound event: an event that consists of more than a single outcome from the sample space

# 13.3 Operations of events using set theory and Venn diagrams

## 13.3.1 Union of events: A U B

Collection of all outcomes that are in A or in B or in both

```{r}
A <- c(2, 4, 6)      # A = {2, 4, 6}
B <- c(1, 2)         # B = {1, 2}

union(A, B)          # A∪B = {2, 4, 6, 1} 
```

```{r}
venn("A + B", zcolor = "#7F7FFF", opacity = 1, ggplot = TRUE) +
  annotate("text", x = 35, y = 950, label = "Ω") +
  annotate("text", x = 500, y = 750, label = "AUB")
```

## 13.3.2 Intersection of events

All outcomes that are in both A and B

```{r}
# A = {2, 4, 6}
# B = {1, 2}

intersect(A, B)
```

```{r}
venn("A B", zcolor = "#7F7FFF", opacity = 1,  ggplot = TRUE) +
  annotate("text", x = 35, y = 950, label = "Ω") +
  annotate("text", x = 500, y = 530, label = "A∩B")
```

## 13.3.3 Complement events

All outcomes of the sample space that don't belong to A U B

```{r}
# A = {2, 4, 6}
# B = {1, 2}

AUB <- union(A, B)                     # A∪B = {2, 4, 6, 1} 


sample_space <- c(1, 2, 3, 4, 5, 6)    # sample_space = {1, 2, 3, 4, 5, 6}

setdiff(sample_space, AUB)
```

```{r}
venn("A + B", zcolor = "white", opacity = 1, box = FALSE, ggplot = TRUE) +
theme(panel.background = element_rect(fill = "#7F7FFF")) +
  annotate("text", x = 5, y = 1000, label = "Ω") +
  annotate("text", x = 500, y = 950, label = "(AUB)^{C}", parse =TRUE)
```

## 13.3.4 Mutually exclusive events

A = "the number rolled is even"

B = "the number rolled is odd"

```{r}
A <- c(2, 4, 6)      
C <- c(1, 3, 5)         

intersect(A, C)         
```

```{r}
# List of items
x2 <- list("A" = A, "C" = C)

ggvenn(x2, fill_alpha = 1, auto_scale = T, show_percentage = FALSE) +
  labs(title = "Ω") +
  scale_fill_manual(values = c("#7F7FFF", "#7F7FFF")) +
  theme(plot.title = element_text(size = 20),
        plot.background = element_rect(fill = "white"))
```

# 13.4 Probability
