---
title: "Normal distribution"
format: html
editor: visual
---

# 15.1 Packages we need

```{r}
library(stevemisc)
library(ggpubr)

library(here)
library(tidyverse)
```

# 15.2 The shape of a normal distribution

```{r}
x <- seq(-4, 4, length=200)                                                    
df <- data.frame(x)                                                            
ggplot(df, aes(x)) +                                                        
  stat_function(fun = dnorm) +                                              
  scale_x_continuous(breaks = c(-3, -2, -1, 0, 1, 2, 3),                    
                     labels = expression(-3*sigma, -2*sigma, -1*sigma,      
                                         mu, 1*sigma, 2*sigma, 3*sigma)) +
  labs(x = "Variable",
       y = "Probability density") +
  theme(text = element_text(size = 16))                                    
```

# 15.3 The properties of a normal distribution

```{r}
normal_dist("#522d80","#00868B") + 
  labs(y = "Probability density", 
       x = "Variable") +
  scale_x_continuous(breaks = c(-3, -2.58, -1.96, -1, 
                              0, 1, 1.96, 2.58, 3),
                   labels = expression(-3*sigma, -2.58*sigma, -1.96*sigma, -1*sigma, 
                                       mu, 1*sigma, 1.96*sigma, 2.58*sigma, 3*sigma)) +
  theme(text = element_text(size = 20), 
        axis.text.x = element_text(size = 12))
```

The normal distribution has the following properties:

-   bell shaped and symmetrical around the mean. Shape statistics, skewness and kurtosis, are zero

-   the peak of the curve lies above the mean

-   any position along the horizontal axis can be expressed as a number of standard deviations from the mean

-   all three measures of central location - mean, median, and mode - are the same

-   the empirical rule (also called the "68-95-99 rule". Much of the area (68%) of the distribution is between -1 $\sigma$ below the mean and +1 $\sigma$ above the mean, the large majority (95%) between -1.96 $\sigma$ below the mean and +1.96 $\sigma$ above the mean (often used as a reference range), and almost all (99%) between -2.58 $\sigma$ below the mean and +2.58 $\sigma$ above the mean. The total area under the curve equals to 1 (100%), almost -3 $\sigma$ below the mean and +3 $\sigma$ above the mean.

# 15.4 Shape statistics and symmetry

## 15.4.1 Skewness

Skewness is usually described as a measure of a distribution's symmetry - or lack of symmetry.

Skewness values that are **negative** indicate a tail to the **left**, **zero** values indicate a **symmetric** distribution, while values that are **positive** indicate a tail to the **right**.

The skewness for a normal distribution is zero. In practice, approximate bell-shaped curves have skewness values between −1 and +1. Values from −1 to −3 or from +1 to +3 indicate that the distribution is tending away from symmetry with \>1 indicating moderate skewness and \>2 indicating severe skewness. Any values **below−3** or **above +3** are a good indication that the distribution is **not symmetric**, therefore, the variable can not be normally distributed.

```{r}
# create a data frame
x <- seq(0, 1, length=200)
y1 <- dbeta(x, 7, 2)
y2 <- dbeta(x, 7, 7)
y3 <- dbeta(x, 2, 7)
df1 <- data.frame(x, y1, y2, y3)
```

-   Left skewed distribution (negatively skewed). The mean and the median are to the left of the mode:

```{r}
# left skewed distribution
ggplot(df1, aes(x, y1)) +
  geom_line(color="green", linewidth = 1.0) +
  geom_segment(aes(x = 0.7, y = 0, xend = 0.7,  yend = 1.98),
               color = "black", linetype = "dashed", linewidth = 0.8) +
  annotate('text', x = 0.67, y = 2.1, label = 'mean', size = 8, color = "black") +
  geom_segment(aes(x = 0.78, y = 0, xend = 0.78,  yend = 2.78), 
               color = "blue", linetype = "dashed", linewidth = 0.8) +
  annotate('text', x = 0.75, y = 2.9, label = 'median', size = 8, color = "blue") +
  geom_segment(aes(x = 0.86, y = 0, xend = 0.86,  yend = 3.17), 
               color = "orange", linetype = "dashed", linewidth = 0.8) +
  annotate('text', x = 0.81, y = 3.13, label = 'mode', size = 8, color = "orange") +
  theme_minimal(base_size = 18) +
  coord_cartesian(expand = FALSE, xlim = c(0, NA), ylim = c(0, NA)) +
  labs(title = "Left skewed distribution",
       x = "Variable",
       y = "Probability density")
```

-   Symmetric distribution (zero skewness). The mean, median and mode are the same

```{r}
# symmetric distribution
ggplot(df1, aes(x, y2)) +
  geom_line(color="green", linewidth = 1.0) +
  geom_segment(aes(x = 0.49, y = 0, xend = 0.49,  yend = 2.89), 
               color = "orange", linetype = "dashed", linewidth = 0.8) +
  annotate('text', x = 0.5, y = 2.46, label = 'mode', size = 8, color = "orange") +
  geom_segment(aes(x = 0.5, y = 0, xend = 0.5,  yend = 2.9), 
               color = "black", linetype = "dashed", linewidth = 0.8) +
  annotate('text', x = 0.5, y = 2.90, label = 'mean', size = 8, color = "black") +
  geom_segment(aes(x = 0.51, y = 0, xend = 0.51,  yend = 2.89), 
               color = "blue", linetype = "dashed", linewidth = 0.8) +
  annotate('text', x = 0.5, y = 2.66, label = 'median', size = 8, color = "blue") +
  theme_minimal(base_size = 18) +
  coord_cartesian(expand = FALSE, xlim = c(0, NA), ylim = c(0, NA)) +
  labs(title = "Symmetric distribution",
       x = "Variable",
       y = "Probability density")
```

-   Right skewed distribution (positively skewed). The mean and median are to the right of the mode

```{r}
# right skewed distribution
ggplot(df1, aes(x, y3)) +
  geom_line(color="green", linewidth = 1.0) +
  geom_segment(aes(x = 0.15, y = 0, xend = 0.15,  yend = 3.17), 
               color = "orange", linetype = "dashed", linewidth = 0.8) +
  annotate('text', x = 0.22, y = 3.10, label = 'mode', size = 8, color = "orange") +
  geom_segment(aes(x = 0.25, y = 0, xend = 0.25,  yend = 2.5), 
               color = "blue", linetype = "dashed", linewidth = 0.8) +
  annotate('text', x = 0.3, y = 2.55, label = 'median', size = 8, color = "blue") +
  geom_segment(aes(x = 0.3, y = 0, xend = 0.3,  yend = 1.9), 
               color = "black", linetype = "dashed", linewidth = 0.8) +
  annotate('text', x = 0.34, y = 1.96, label = 'mean', size = 8, color = "black") +
  theme_minimal(base_size = 18) +
  coord_cartesian(expand = FALSE, xlim = c(0, NA), ylim = c(0, NA)) +
  labs(title = "Right skewed distribution",
       x = "Variable",
       y = "Probability density")
```

## 15.4.2 Excess kurtosis

The other way that distributions can deviate from normality is kurtosis. The **excess kurtosis** parameter is a measure of the combined weight of the tails relative to the rest of the distribution.

Kurtosis is associated indirectly with the peak of the distribution (if the peak of the distribution is too high/sharp or too low compared to a "normal" distribution).

Distributions with **negative** kurtosis are called **platykurtic**.

If the measure of kurtosis is **zero** the distribution is **mesokurtic**.

Finally, distributions with **positive** excess kurtosis are called **leptokurtic**.

The excess kurtosis for a normal distribution is zero. In practice, approximate normal distributions have excess kurtosis values between −1 and +1. Values from −1 to −3 or from +1 to +3 indicate that the distribution is tending away from a mesokurtic distribution. Any values **below −3** or **above +3** are a good indication that the distribution is **not mesokurtic**, therefore, the variable can not be normally distributed.

```{r}
# create a data frame
x <- seq(-6, 6, length=200)
y1 <- dnorm(x)
y2 <- dnorm(x, sd= 2)
y3 <- dnorm(x, sd= 0.5)
df2 <- data.frame(x, y1, y2, y3)
```

-   Platykurtic distribution (negative excess kurtosis)

```{r}
# platykurtic distribution
ggplot(df2, aes(x, y1)) +
  geom_line(color="black", linewidth = 0.8, linetype = "dashed") +
  geom_line(aes(x, y2), color="deeppink", linewidth = 0.8) +
  annotate('text', x = 2.0, y = 0.3, label = 'normal', size = 7, color = "black") +
  annotate('text', x = 4.2, y = 0.1, label = 'platykurtic', size = 8, color = "deeppink") +
  theme_minimal(base_size = 18) +
  labs(title = "Platykurtic distribution",
       x = "Variable",
       y = "Probability density")
```

-   Mesokurtic distribution (zero excess kurtosis)

```{r}
# mesokurtic distribution
ggplot(df2, aes(x, y1)) +
  geom_line(color="deeppink", linewidth = 0.8) +
  annotate('text', x = 0, y = 0.20, label = 'normal is a', size = 7, color = "black") +
  theme_minimal(base_size = 18) +
  annotate('text', x = 0, y = 0.15, label = 'mesokurtic distribution', size = 7, color = "deeppink") +
  labs(title = "Mesokurtic distribution",
       x = "Variable",
       y = "Probability density")
```

-   Leptokurtic distribution (positive excess kurtosis)

```{r}
# leptokurtic distribution
ggplot(df2, aes(x, y1)) +
  geom_line(color="black", linewidth = 0.8, linetype = "dashed") +
  geom_line(aes(x, y3), color="deeppink", linewidth = 0.8) +
  annotate('text', x = 2.1, y = 0.25, label = 'normal', size = 7, color = "black") +
  annotate('text', x = 1.9, y = 0.75, label = 'leptokurtic', size = 8, color = "deeppink") +
  theme_minimal(base_size = 18) +
  labs(title = "Leptokurtic distribution",
       x = "Variable",
       y = "Probability density")
```

# 15.5 Normal Q-Q plots

The normal Q-Q plot, or normal quantile-quantile plot, provides an easy way to visually check whether or not a data set is normally distributed.

The values in the plot are the **quantiles** of the variable distribution (sample quantiles) plotted against the quantiles of a standard normal distribution (theoretical quantiles).

If the points fall close to a **straight line** at a 45-degree angle, then the data are normally distributed (although the ends of the Q-Q plot often deviate from the straight line).

-   Normal distribution

```{r}
set.seed(48)
x1 <- rnorm(1000, 20, 0.05)
x2 <- rgamma(1000, shape = 2, rate = 2)
df3 <- data.frame(x1, x2)

# density plot of age
ggplot(df3, aes(x = x1)) +
  geom_density(fill="steelblue4", color="#8fb4d9", 
               adjust = 1.5, alpha=0.6) +
  theme_minimal(base_size = 20)
```

```{r}
qqnorm(df3$x1, pch = 1, frame = FALSE)
qqline(df3$x1, col = "steelblue", lwd = 2)
```

```{r}
ggqqplot(df3, "x1", conf.int = F) +
  stat_qq_line(color="blue")
```

-   Skewed distribution

```{r}
# density plot of age
ggplot(df3, aes(x = x2)) +
  geom_density(fill="steelblue4", color="#8fb4d9", 
               adjust = 1.5, alpha=0.6) +
  theme_minimal(base_size = 20)
```

```{r}
qqnorm(df3$x2, pch = 1, frame = FALSE)
qqline(df3$x2, col = "steelblue", lwd = 2)
```

```{r}
ggqqplot(df3, "x2", conf.int = F) +
  stat_qq_line(color="blue")
```
